\bf Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-03-11 paper.qmd
Add Methodology with mathematical expressions

**Step1: Cluster 2D points to hexagons**

Expanding upon the information regarding hexagonal binning discussed in Step 1 of @sec-construct2d, the primary objective in this step is to determine the 2D embedding points associated with each hexagon. As the initial process of hexagonal binning, the 2D points are clustered into their respective hexagonal bins. By mapping this information with the hexagonal bin centroids ($C_k^{(2)} \equiv (C_{ky_1}, C_{ky_2})$) that obtained in Step 2 of the 2D model building (see @sec-construct2d), we can find which 2D points are assigned to which data set (see @fig-wkhighD (a)).

**Step2: Cluster high dimensional points to hexagons**

Following the step 1, the main focus in this step to determine the corresponding high dimensional points for each hexagon. Every 2D embedding point serves as a projection of a data point belonging to high dimensional space. By using this mapping between the high dimensional data and their corresponding projections, the high dimensional points allocated to each hexagons are determined (see video linked in @fig-wkhighD).

**Step3: Compute the mean within hexagon**

Having identified the high-dimensional points associated with hexagons, the final step involves computing the mean within each hexagonal bin. This implies calculating the average of the high-dimensional data points located within each hex bin. These averaged high-dimensional data points, denoted as $C_k^{(p)} \equiv (C_{kx_1}, ..., C_{kx_p})$, serve as the representative coordinates for the hex bin centroids within the expansive high-dimensional space (see video linked in @fig-wkhighD).

@@

Based on the insights gained in Step 3 of @sec-construct2d, where connected edges in the 2D triangular mesh were identified, this step involves lifting these connections into the high dimensions. Having the mappings of all 2D triangular mesh points into high dimensions, the points connected in 2D are also connected in high dimensional space (see video linked in @fig-wkhighD).

@@

## To obtain the standardise counts within hexbins
counts_df <- as.data.frame(do.call(cbind, hb_obj_tsne_gau$std_cts))
df_bin_centroids <- extract_hexbin_centroids(centroids_df = all_centroids_df, 
                                             counts_df = counts_df)
ggplot(data = hex_grid, aes(x = x, y = y)) + 
  geom_polygon(fill = "white", color = "black", aes(group = hex_poly_id)) +
  geom_point(data = all_centroids_df, aes(x = c_x, y = c_y), color = "red") +
  coord_fixed()

@@

ggplot(data = hex_grid_with_counts, aes(x = x, y = y)) +
  geom_polygon(color = "black", aes(group = hex_poly_id, fill = std_counts)) +
  geom_text(data = all_centroids_df, aes(x = c_x, y = c_y, label = hexID)) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff") +
  coord_fixed()

@@

distance_plot <- plot_dist(distance_tsne_gau) +
  ylab(expression(d^{(2)})) +
  theme(axis.text = element_text(size = 5),
        axis.title = element_text(size = 12))

distance_plot

@@

ggplot() +
geom_trimesh(data = df_bin_centroids_tsne_gau, mapping = aes(x = c_x, y = c_y))

@@

## To obtain the standardise counts within hexbins
counts_df <- as.data.frame(do.call(cbind, hb_obj_pbmc$std_cts))
df_bin_centroids <- extract_hexbin_centroids(centroids_df = all_centroids_df, 
                                             counts_df = counts_df)
ggplot(data = hex_grid, aes(x = x, y = y)) + 
  geom_polygon(fill = "white", color = "black", aes(group = hex_poly_id)) +
  geom_point(data = all_centroids_df, aes(x = c_x, y = c_y), color = "red") +
  coord_fixed()

@@

distance_plot <- plot_dist(distance_pbmc) +
  ylab(expression(d^{(2)})) +
  theme(axis.text = element_text(size = 5),
        axis.title = element_text(size = 12))

distance_plot

@@

ggplot() +
geom_trimesh(data = df_bin_centroids_pbmc, mapping = aes(x = c_x, y = c_y))

@@

ggplot(data = hex_grid, aes(x = x, y = y)) + 
  geom_polygon(fill = "white", color = "black", aes(group = hex_poly_id)) +
  geom_point(data = all_centroids_df, aes(x = c_x, y = c_y), color = "red") +
  coord_fixed()

>>>>

Consider $f: \mathbb{R}^p \rightarrow \mathbb{R}^2$ be a function that maps the high-D data ($X_{n \times p}$) to its NLDR equivalent ($Y_{n \times d}$). Then, let $g: \mathbb{R}^2 \rightarrow \mathbb{R}^2$ be a function that maps each 2D embedding point to its closest centroid ($C^{(2)}$). It follows that $f(g(x))$ maps the high-D points $x$ to the centroid in 2D ($C_k^{(2)}$). Also, define a function $v: \mathbb{R}^2 \rightarrow \mathbb{R}^p$ maps the 2D centroid ($C^{(2)}$) to the high-D mean of the points ($C^{(p)}$) in the hexagon.

The high-D mean of all the points in $k^{th}$ hexagon by

$$
C_k^{(p)} = (C_{kx_1}, ..., C_{kx_p}) = \left(\frac{1}{n_k} \sum_{i=1}^{n_k} x_{1i}, \frac{1}{n_k} \sum_{i=1}^{n_k} x_{2i}, \dots ,\frac{1}{n_k} \sum_{i=1}^{n_k} x_{pi}\right).
$${#eq-equation6}

Therefore,

$$
v(C_k^{(2)}) = C_k^{(p)}.
$${#eq-equation7}

Therefore, $f(g(x))$ gives the 2D centroid associated with high-D points $x$, and $v(C_k^{(2)})$ gives the high-D centroid associated with 2D point $C_k^{(2)}$. Thus, $v(f(g(x)))$ gives the high-D centroid ($C_k^{(p)}$) associated with the 2D embedding of the points $x$.

@@

As described in Step 4 of @sec-construct2d, during the triangulation process in 2D space, vertices are identified to form edges. With the knowledge of the high-D mappings for the 2D hexagonal bins, the vertices connected in 2D are also connected in high-D (see video linked in @fig-wkhighD).

@@

## To obtain the standardise counts within hexbins
counts_df <- as.data.frame(do.call(cbind, hb_obj_tsne_gau$std_cts))
df_bin_centroids <- extract_hexbin_centroids(centroids_df = all_centroids_df, 
                                             counts_df = counts_df)
# ggplot(data = hex_grid, aes(x = x, y = y)) + 
#   geom_polygon(fill = "white", color = "black", aes(group = hex_poly_id)) +
#   geom_point(data = all_centroids_df, aes(x = c_x, y = c_y), color = "red") +
#   coord_fixed()

@@

@@

# distance_plot <- plot_dist(distance_tsne_gau) +
#   ylab(expression(d^{(2)})) +
#   theme(axis.text = element_text(size = 5),
#         axis.title = element_text(size = 12))
# 
# distance_plot

@@

# ggplot() +
# geom_trimesh(data = df_bin_centroids_tsne_gau, mapping = aes(x = c_x, y = c_y))

@@

## To obtain the standardise counts within hexbins
counts_df <- as.data.frame(do.call(cbind, hb_obj_pbmc$std_cts))
df_bin_centroids <- extract_hexbin_centroids(centroids_df = all_centroids_df, 
                                             counts_df = counts_df)
# ggplot(data = hex_grid, aes(x = x, y = y)) + 
#   geom_polygon(fill = "white", color = "black", aes(group = hex_poly_id)) +
#   geom_point(data = all_centroids_df, aes(x = c_x, y = c_y), color = "red") +
#   coord_fixed()

@@

# distance_plot <- plot_dist(distance_pbmc) +
#   ylab(expression(d^{(2)})) +
#   theme(axis.text = element_text(size = 5),
#         axis.title = element_text(size = 12))
# 
# distance_plot

@@

# ggplot() +
# geom_trimesh(data = df_bin_centroids_pbmc, mapping = aes(x = c_x, y = c_y))

@@

# ggplot(data = hex_grid, aes(x = x, y = y)) + 
#   geom_polygon(fill = "white", color = "black", aes(group = hex_poly_id)) +
#   geom_point(data = all_centroids_df, aes(x = c_x, y = c_y), color = "red") +
#   coord_fixed()

### 2024-03-11 paper.tex
Add Methodology with mathematical expressions

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, enhanced, sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, interior hidden, breakable]}{\end{tcolorbox}}\fi

@@

\textbf{Step1: Cluster 2D points to hexagons}

Expanding upon the information regarding hexagonal binning discussed in
Step 1 of Section~\ref{sec-construct2d}, the primary objective in this
step is to determine the 2D embedding points associated with each
hexagon. As the initial process of hexagonal binning, the 2D points are
clustered into their respective hexagonal bins. By mapping this
information with the hexagonal bin centroids
(\(C_k^{(2)} \equiv (C_{ky_1}, C_{ky_2})\)) that obtained in Step 2 of
the 2D model building (see Section~\ref{sec-construct2d}), we can find
which 2D points are assigned to which data set (see
\textbf{?@fig-wkhighD} (a)).

\textbf{Step2: Cluster high dimensional points to hexagons}

Following the step 1, the main focus in this step to determine the
corresponding high dimensional points for each hexagon. Every 2D
embedding point serves as a projection of a data point belonging to high
dimensional space. By using this mapping between the high dimensional
data and their corresponding projections, the high dimensional points
allocated to each hexagons are determined (see video linked in
\textbf{?@fig-wkhighD}).

\textbf{Step3: Compute the mean within hexagon}

Having identified the high-dimensional points associated with hexagons,
the final step involves computing the mean within each hexagonal bin.
This implies calculating the average of the high-dimensional data points
located within each hex bin. These averaged high-dimensional data
points, denoted as \(C_k^{(p)} \equiv (C_{kx_1}, ..., C_{kx_p})\), serve
as the representative coordinates for the hex bin centroids within the
expansive high-dimensional space (see video linked in
\textbf{?@fig-wkhighD}).

@@

Based on the insights gained in Step 3 of Section~\ref{sec-construct2d},
where connected edges in the 2D triangular mesh were identified, this
step involves lifting these connections into the high dimensions. Having
the mappings of all 2D triangular mesh points into high dimensions, the
points connected in 2D are also connected in high dimensional space (see
video linked in \textbf{?@fig-wkhighD}).

@@

\includegraphics{paper_files/figure-pdf/unnamed-chunk-44-1.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-44-2.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-44-3.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-44-4.pdf}

@@

\includegraphics{paper_files/figure-pdf/unnamed-chunk-59-1.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-59-2.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-59-3.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-59-4.pdf}

@@

\includegraphics{paper_files/figure-pdf/unnamed-chunk-65-1.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-65-2.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-65-3.pdf}

\includegraphics{paper_files/figure-pdf/unnamed-chunk-65-4.pdf}

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, breakable, boxrule=0pt, enhanced, borderline west={3pt}{0pt}{shadecolor}, interior hidden, sharp corners]}{\end{tcolorbox}}\fi

@@

Consider \(f: \mathbb{R}^p \rightarrow \mathbb{R}^2\) be a function that
maps the high-D data (\(X_{n \times p}\)) to its NLDR equivalent
(\(Y_{n \times d}\)). Then, let
\(g: \mathbb{R}^2 \rightarrow \mathbb{R}^2\) be a function that maps
each 2D embedding point to its closest centroid (\(C^{(2)}\)). It
follows that \(f(g(x))\) maps the high-D points \(x\) to the centroid in
2D (\(C_k^{(2)}\)). Also, define a function
\(v: \mathbb{R}^2 \rightarrow \mathbb{R}^p\) maps the 2D centroid
(\(C^{(2)}\)) to the high-D mean of the points (\(C^{(p)}\)) in the
hexagon.

The high-D mean of all the points in \(k^{th}\) hexagon by

\begin{equation}\protect\hypertarget{eq-equation6}{}{
C_k^{(p)} = (C_{kx_1}, ..., C_{kx_p}) = \left(\frac{1}{n_k} \sum_{i=1}^{n_k} x_{1i}, \frac{1}{n_k} \sum_{i=1}^{n_k} x_{2i}, \dots ,\frac{1}{n_k} \sum_{i=1}^{n_k} x_{pi}\right).
}\label{eq-equation6}\end{equation}

Therefore,

\begin{equation}\protect\hypertarget{eq-equation7}{}{
v(C_k^{(2)}) = C_k^{(p)}.
}\label{eq-equation7}\end{equation}

Therefore, \(f(g(x))\) gives the 2D centroid associated with high-D
points \(x\), and \(v(C_k^{(2)})\) gives the high-D centroid associated
with 2D point \(C_k^{(2)}\). Thus, \(v(f(g(x)))\) gives the high-D
centroid (\(C_k^{(p)}\)) associated with the 2D embedding of the points
\(x\).

@@

As described in Step 4 of Section~\ref{sec-construct2d}, during the
triangulation process in 2D space, vertices are identified to form
edges. With the knowledge of the high-D mappings for the 2D hexagonal
bins, the vertices connected in 2D are also connected in high-D (see
video linked in \textbf{?@fig-wkhighD}).

@@

@@

@@
